Mechanical loading prediction through accelerometry data during walking and running

### 2021-09-24 manuscript/manuscript.Rmd
Do some small corrections in the text

Force plates (FP) are the golden standard instrument for mechanical loading measurement [@Medved2000]. However, their cost, need of expertise to collect and analyze data, of a dedicated indoor space and ongoing calibration and maintenance hinders its application in free-living conditions. Accelerometers, due to their portability, low-cost and reliability, are widely used to objectively measure physical activity related to cardiometabolic outcomes [@Troiano2014; @Migueles2017], and, more recently, have been proposed as adequate devices to predict gravitational loading variables [@Neugebauer2012; @Fortune2014; @Neugebauer2014; @Neugebauer2018; @Veras2020]. Given the need to assess mechanical loading in out-of-laboratory environments to control and monitor exercise prescription for bone health, accurate mechanical loading prediction models need to be developed and validated for the most common physical activities prescribed to induce bone adaptations, such as walking, running, and jumping. These models, apart from being accurate, need to be transparent in their data processing methods, allowing reproducibility of the results, and ideally offer a tool to facilitate data processing and analysis for those who are not familiar with the employed techniques, such as what is already offered for analysing cardiometabolic variables [@Migueles2019], and therefore increase their applicability.

@@

Regression equations were developed to predict the mechanical loading variables based on accelerometry data. In the pGRF prediction equations, body mass and pACC were significant predictors and, therefore, included in the models. Table 1 shows the regression coefficients, R^2^ and accuracy indices from models derived from accelerometers placed at ankle, lower back and hip, and from the resultant vector and its vertical component. In the pGRF prediction models, the lowest R^2^ value found was `r min(GRF_accuracy$R2_cond)`, which shows that the predictor variables are able to explain the variability in the outcome variable to a great extent. Also, MAE ranged between `r myround(min(GRF_accuracy$MAE), 1)` N and `r myround(max(GRF_accuracy$MAE), 1)` N, and the MAPE from `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% to `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which is a smaller error than in previously published equations [@Neugebauer2014; @Fortune2014], and similar to the error from equations previously published by our group for walking only [@Veras2020]. RMSE was also computed and shown to range between `r myround(min(GRF_accuracy$RMSE), 1)` N and `r myround(max(GRF_accuracy$RMSE), 1)` N. Although all pGRF prediction equations showed very similar results, the one based on hip-worn accelerometers and the resultant vector has shown the best performance, with an R^2^ of `r hip_GRF_accuracy$R2_cond`, the lowest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

@@

Comparing our results with previous studies that used a similar approach to predict pGRF, it can be seen that prediction errors from our models are comparable with some studies whose best results presented a MAPE of approximately 6.0\%, but which included only walking in their protocol [@Fortune2014; @Neugebauer2018; @Veras2020]. Additionally, previous studies that developed pGRF prediction models during both walking and running showed a MAPE of 8.3\% [@Neugebauer2014], which is higher than our results, and 5.2\% [@Neugebauer2012] with no model cross-validation, which is similar to our results when no cross-validation method is applied (highest value of `r accuracy_no_cv`\% in our models). Nevertheless, models from both studies present a caveat, that is the inclusion of a term in their equations that distinguishes whether the activity is walking or running, which is not easy to be identified from free-living accelerometer data, hindering the real-world application of these previous equations. Also, one of these studies included in their sample only children and adolescents [@Neugebauer2012], which present a higher variability in their gait pattern compared to the adults [@Gouelle2016], making this equation unsuitable to use in an adult population. Furthermore, it was observed that the reference equation [@Neugebauer2014] used as reference has a much lower accuracy during walking compared with running (16.6\% vs. 9.7\%). As this reference equation was not validated in a sample including people with obesity, it performs poorly when applied to data from people with high body mass, as is the case in our sample that performed the walking trials, compared with running trials mostly performed by non-obese people. It is also important to mention that the prediction error found in our pGRF models is lower than the error typically found in models developed to estimate energy expenditure [@Diniz-Sousa2020], which is one of the main uses of accelerometer data [@Migueles2017].

@@

Our findings revealed that ankle, lower back or hip were all valid accelerometer placements to accurately predict pGRF. However, despite the similarities, hip was the accelerometer placement with the best prediction accuracy, which reinforces it as the preferable location to collect and analyse accelerometry-measured physical activity data. This placement is already well accepted for energy expenditure prediction and physical activity intensity classification [@Migueles2017] and has also shown a high mechanical loading prediction ability in previous studies [@Veras2020; @Neugebauer2018; @Neugebauer2012]. Besides various placements, the current study also analysed predictions from both the resultant and vertical vectors. Although ACC and GRF from the vertical vector are the most studied and frequently associated with bone health [@Vainionpaa2006; @Heikkinen2007], the resultant vector of both ACC and GRF during locomotion is greatly determined by the vertical vector, and, as can be observed in Supplemental Figure S1, in this study values from both vectors were very similar. Also, mechanical loading prediction based on the resultant ACC offers the important advantage of not being dependent on the accelerometer positioning and correct orientation, which is hard to assure when measuring daily physical activity [@Welk2005; @Wundersitz2015].

@@

Besides these advantages, our models present some important potential clinical applications. For instance, these models can be used to analyse data from large observational studies and make associations between mechanical loading and bone health. But instead of analysing accelerometer data averaged in epochs similarly to some previous publications [@Chastin2014; @Stiles2017], which can dilute the higher intensities [@Fabre2020], with our method raw ACC can be analysed at a higher resolution by identifying the pGRF associated with each step taken and establishing, for example, the number of peaks above a certain magnitude threshold needed to elicit a positive bone adaptation. Another potential application could be the computation of the cumulative mechanical loading by adding up the pGRF of all detected steps [@Diniz-Sousa2021]. This information could be used either for the determination of the relationship between the mechanical loading accumulated in a certain period (e.g., a day) and the bone response to this loading [@Ahola2010], or to investigate the causes of bone loss in populations with bone health issues, such as bariatric patients [@Diniz-Sousa2021]. This approach has already been applied with equations developed in a previous research from our group [@Veras2020], but now can be extended to include both types of locomotion (walking and running) contemplated in the equations from the current study. Another possible application is in sports training, by monitoring the amount of loading beared by athletes and relate this information with osteoarticular safety of the training regimen [@Vanwanseele2020].

>>>>

Force plates (FP) are the golden standard instrument for mechanical loading measurement [@Medved2000]. However, their cost, need of expertise to collect and analyze data, of a dedicated indoor space and ongoing calibration and maintenance hinders its application in free-living conditions. Accelerometers, due to their portability, low-cost and reliability, are widely used to objectively measure physical activity related to cardiometabolic outcomes [@Troiano2014; @Migueles2017], and, more recently, have been proposed as adequate devices to predict gravitational loading variables [@Neugebauer2012; @Fortune2014; @Neugebauer2014; @Neugebauer2018; @Veras2020]. Given the need to assess mechanical loading in out-of-laboratory environments to control and monitor exercise prescription for bone health, accurate mechanical loading prediction models need to be developed and validated for the most common physical activities prescribed to induce bone adaptations, such as walking, running, and jumping. These models, apart from being accurate, need to be transparent in their data processing methods, allowing reproducibility of the results, and ideally offer a tool to facilitate data processing and analysis for those who are not familiar with the employed techniques, such as what is already offered for analyzing cardiometabolic variables [@Migueles2019], and therefore increase their applicability.

@@

Regression equations were developed to predict the mechanical loading variables based on accelerometry data. In the pGRF prediction equations, body mass and pACC were significant predictors and, therefore, included in the models. Table 1 shows the regression coefficients, R^2^ and accuracy indices from models derived from accelerometers placed at ankle, lower back and hip, and from the resultant vector and its vertical component. In the pGRF prediction models, the lowest R^2^ value found was `r min(GRF_accuracy$R2_cond)`, which shows that the predictor variables can explain the variability in the outcome variable to a great extent. Also, MAE ranged between `r myround(min(GRF_accuracy$MAE), 1)` N and `r myround(max(GRF_accuracy$MAE), 1)` N, and the MAPE from `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% to `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which is a smaller error than in previously published equations [@Neugebauer2014; @Fortune2014], and similar to the error from equations previously published by our group for walking only [@Veras2020]. RMSE was also computed and shown to range between `r myround(min(GRF_accuracy$RMSE), 1)` N and `r myround(max(GRF_accuracy$RMSE), 1)` N. Although all pGRF prediction equations showed very similar results, the one based on hip-worn accelerometers and the resultant vector has shown the best performance, with an R^2^ of `r hip_GRF_accuracy$R2_cond`, the lowest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

@@

Comparing our results with previous studies that used a similar approach to predict pGRF, prediction errors from our models are comparable with some studies whose best results presented a MAPE of approximately 6.0\%, but which included only walking in their protocol [@Fortune2014; @Neugebauer2018; @Veras2020]. Additionally, previous studies that developed pGRF prediction models during both walking and running showed a MAPE of 8.3\% [@Neugebauer2014], which is higher than our results, and 5.2\% [@Neugebauer2012] with no model cross-validation, which is similar to our results when no cross-validation method is applied (highest value of `r accuracy_no_cv`\% in our models). Nevertheless, models from both studies present a caveat, that is the inclusion of a term in their equations that distinguishes whether the activity is walking or running, which is not easy to be identified from free-living accelerometer data, hindering the real-world application of these previous equations. Also, one of these studies included in their sample only children and adolescents [@Neugebauer2012], which present a higher variability in their gait pattern compared to the adults [@Gouelle2016], making this equation unsuitable to use in an adult population. Furthermore, it was observed that the reference equation [@Neugebauer2014] used as reference has a much lower accuracy during walking compared with running (16.6\% vs. 9.7\%). As this reference equation was not validated in a sample including people with obesity, it performs poorly when applied to data from people with high body mass, as is the case in our sample that performed the walking trials, compared with running trials mostly performed by non-obese people. It is also important to mention that the prediction error found in our pGRF models is lower than the error typically found in models developed to estimate energy expenditure [@Diniz-Sousa2020], which is one of the main uses of accelerometer data [@Migueles2017].

@@

Our findings revealed that ankle, lower back or hip were all valid accelerometer placements to accurately predict pGRF. However, despite the similarities, hip was the accelerometer placement with the best prediction accuracy, which reinforces it as the preferable location to collect and analyze accelerometry-measured physical activity data. This placement is already well accepted for energy expenditure prediction and physical activity intensity classification [@Migueles2017] and has also shown a high mechanical loading prediction ability in previous studies [@Veras2020; @Neugebauer2018; @Neugebauer2012]. Besides various placements, the current study also analyzed predictions from both the resultant and vertical vectors. Although ACC and GRF from the vertical vector are the most studied and frequently associated with bone health [@Vainionpaa2006; @Heikkinen2007], the resultant vector of both ACC and GRF during locomotion is greatly determined by the vertical vector, and, as can be observed in Supplemental Figure S1, in this study values from both vectors were very similar. Also, mechanical loading prediction based on the resultant ACC offers the important advantage of not being dependent on the accelerometer positioning and correct orientation, which is hard to assure when measuring daily physical activity [@Welk2005; @Wundersitz2015].

@@

Besides these advantages, our models present some important potential clinical applications. For instance, these models can be used to analyze data from large observational studies and make associations between mechanical loading and bone health. But instead of analyzing accelerometer data averaged in epochs similarly to some previous publications [@Chastin2014; @Stiles2017], which can dilute the higher intensities [@Fabre2020], with our method raw ACC can be analyzed at a higher resolution by identifying the pGRF associated with each step taken and establishing, for example, the number of peaks above a certain magnitude threshold needed to elicit a positive bone adaptation. Another potential application could be the computation of the cumulative mechanical loading by adding up the pGRF of all detected steps [@Diniz-Sousa2021]. This information could be used either for the determination of the relationship between the mechanical loading accumulated in a certain period (e.g., a day) and the bone response to this loading [@Ahola2010], or to investigate the causes of bone loss in populations with bone health issues, such as bariatric patients [@Diniz-Sousa2021]. This approach has already been applied with equations developed in a previous research from our group [@Veras2020], but now can be extended to include both types of locomotion (walking and running) contemplated in the equations from the current study. Another possible application is in sports training, by monitoring the amount of loading beared by athletes and relate this information with osteoarticular safety of the training regimen [@Vanwanseele2020].
